Parsing Postfix log files

### 2007-04-16 doc/logparser.tex
Improved the tracking section, and extended it with a description of the algorithm. Changed faked to "of unknown origin" to make it clearer. Removed unnecessary action references cluttering up the algorithm description. Rewrote some of the algorithm stuff so it's hopefully clearer.

If the database supports sub-selects percentages can be
obtained:\footnote{This works for SQLite3, where || is the concatenation
operator.  If the database containing the data does not support
concatenation, simply remove ``|| '\%\''' from the query.}

@@

        Because of this the parser cannot complain when it encounters a log
        line from qmgr for a previously unseen mail; it must mark the mail
        as potentially fake and subsequently clear the faked flag if and
        when the origin of the mail becomes clear.  Obviously the parser
        could omit checking of where mails originate from, but I feel that
        it is better to require an explicit source, as bugs are more likely
        to be exposed.

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails marked as faked cannot be entered into the database when
        their final log line is parsed; instead they must be marked as
        database ready and subsequently entered by the parent mail once it
        has been identified.  Quite apart from identifying mail injected in
        an unknown fashion, or bugs in the parser, faked mails need to be
        marked as such because they lack some data present in the parent
        mail, and must copy that data from their parent before being
        entered in the database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND
        ON IT SOME MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.

@@

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The flow chart in section~\ref{flow-chart} should
be consulted while reading this section also.

@@

    \item The client disconnects (action: disconnect; flowchart:~6).  As
        soon as a mail is accepted it is assigned a queueid; if there is no
        queueid Postfix didn't accept any mail over that connection, it
        rejected all attempted deliveries due to the configured
        restrictions.  Absence of a queueid is checked for during
        disconnection handling (action: still disconnect; flowchart:~7) and
        if none is found the data is cleaned up and entered in the database
        (action: still disconnect; flowchart:~8; subroutines:
        fixup\_connection and commit\_connection), then removed from memory
        (action: still disconnect; flowchart:~9).

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~11) --- see
        section~\ref{tracking-re-injected-mail}.

    \item Eventually the mail will be delivered (even if re-injected),
        bounced, or deleted by the administrator (action: commit;
        flowchart:~12).  This is the last log line for this particular mail
        (though it may be indirectly referred to if it was re-injected).
        If it is neither parent nor child of re-injection the data is
        cleaned up and entered in the database (action: still commit;
        flowchart:~14; subroutines fixup\_connection and
        commit\_connection), then deleted from memory (action: still
        commit; flowchart:~15).  For re-injected mails see
        section~\ref{tracking-re-injected-mail}.

@@

I'd like to reiterate that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is when the mail is
re-injected due to forwarding, and that deserves its own section, which is
coming up next.

@@

It's probably apparent by now that tracking re-injected mails is the single
most complex part of the parser.  Section~\ref{complications} complication
3 has most of the previous discussion, should you wish to review it.

XXX EXPAND THIS

there are two implicit indications:

@@

    \item The more commonly occurring indicator is qmgr selecting a mail
        with a previously unseen queueid for delivery (action:
        qmgr\_chooses\_mail; flowchart:~3), in which case a new data
        structure will be created.  The mail will be marked as faked; this
        flag should be subsequently cleared.  The data is then saved
        (action: still qmgr\_chooses\_mail; flowchart:~3).

    \item This indicator will be seen for every mail that is re-injected
        but is less common in identifying re-injected mail because it
        typically occurs after the indicator explained above.  Local
        delivery re-injects the mail and logs a successful delivery rather
        than delivering directly to a mailbox or program (action: track;
        flowchart:~11).  In this case the mail may already have been
        created (by qmgr\_chooses\_mail, above); if not a new data
        structure will be created.  In both cases the new mail is marked as
        a child of the parent.

@@

>>>>

If the database supports sub-selects percentages can be
obtained:\footnote{This works for SQLite3, where || is the concatenation
operator.  If the database containing the data does not support
this syntax, simply remove ``|| '\%\''' from the query; the results will be
the same, just slightly less visually pleasing.}

@@

        Because of this the parser cannot complain when it encounters a log
        line from qmgr for a previously unseen mail; it must flag the mail
        as coming from an unknown origin and subsequently clear the flag if and
        when the origin of the mail becomes clear.  Obviously the parser
        could omit checking of where mails originate from, but I feel that
        it is better to require an explicit source, as bugs are more likely
        to be exposed.

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails flagged as coming from an unknown origin cannot be
        entered into the database when their final log line is parsed;
        instead they must be marked as database ready and subsequently
        entered by the parent mail once it has been identified.  Quite
        apart from identifying mail injected in an unknown fashion, or bugs
        in the parser, unknown origin mails need to be marked as such
        because they lack some data present in the parent mail, and must
        copy that data from their parent before being entered in the
        database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND ON IT SOME
        MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.

@@

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The data flow chart in section~\ref{flow-chart}
should be consulted while reading this section also.

@@

    \item The client disconnects (action: disconnect; flowchart:~6).  As
        soon as a mail is accepted it is assigned a queueid; thus if there
        is no queueid Postfix didn't accept any mail over that connection,
        it rejected all attempted deliveries due to the configured
        restrictions.  Absence of a queueid is checked for during
        disconnection handling (flowchart:~7) and if none is found the data
        is cleaned up and entered in the database (flowchart:~8;
        subroutines: fixup\_connection and commit\_connection), then
        removed from memory (flowchart:~9).

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~11) --- see
        section~\ref{tracking-re-injected-mail} for further details.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14; subroutines fixup\_connection and
        commit\_connection), then deleted from memory (flowchart:~15).  For
        re-injected mails see section~\ref{tracking-re-injected-mail}.

@@

I'd like to reiterate that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, and that
is explained in section~\ref{tracking-re-injected-mail}.

@@

It's probably apparent by now that tracking re-injected mails is the single
most complex part of the parser.  Section~\ref{complications}, complication
3 has most of the previous discussion, should you wish to review it.

The crux of the problem is that re-injected mails appear in the logs
without explicit logging indicating their source.  There are two implicit
indications:

@@

    \item The indicator which more commonly introduces re-injection is qmgr
        selecting a mail with a previously unseen queueid for delivery
        (action: qmgr\_chooses\_mail; flowchart:~3), in which case a new
        data structure will be created.  The mail will be flagged as having
        unknown origins; this flag should be subsequently cleared once the
        origin has been established.  The data is then saved (flowchart:~3;
        subroutine: save).

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would (action: track; flowchart:~11).\footnote{Relayed
        delivery is performed by the SMTP client; local delivery means
        local to the server, i.e.\ an address the server accepts mail for.}
        In this case the mail may already have been created (action:
        qmgr\_chooses\_mail; flowchart:~3; described above) and the unknown
        origin flag will be cleared; if not a new data structure will be
        created.  In both cases the new mail is marked as a child of the
        parent.  The log line in question is:

        This second indicator always occurs for re-injected mail but
        typically it occurs after the first indicator explained above, and
        so rarely introduces re-injection.  This indicator is required to
        tie the parent and child mails together and so is central to the
        process of tracking re-injected mails.

@@

The algorithm for tracking and saving re-injected mail to the database will
finally be described:

    \item If the mail is of unknown origin it is a child mail whose parent
        hasn't yet been identified (action: commit; flowchart:~16).  Mark
        the mail as ready for entry in the database (flowchart:~17), and
        wait for the parent to deal with it ( flowchart:~18).  The mail
        should never have another directly associated log entry; only its
        parent should refer to it.

    \item If the mail is a child mail then it has already been tracked
        (action: commit; flowchart:~19): the data is cleaned up, pulling
        data from its parent if necessary, and entered in the database
        (flowchart:~20; subroutines fixup\_connection and
        commit\_connection), then deleted from memory (flowchart:~21).  The
        child mail will inform the parent mail that it is finished with
        (flowchart:~22); if this is the last child and the parent is also
        finished the parent will delete itself.  The child mail is now
        finished with.

    \item The last alternative is that the mail is a parent mail (action:
        commit; flowchart:~23).  Regardless of the state of its children
        the data is cleaned up and entered in the database ( flowchart:~24;
        subroutines fixup\_connection and commit\_connection).  The parent
        may have children which are waiting to be entered in the database (
        flowchart:~25); for each of those children their data is cleaned up
        and entered in the database, then the child is finished with and
        its data discarded.  The parent may also have outstanding children
        which are not yet delivered, in which case (flowchart:~26) the
        parent must wait for all children to finish.  As soon as the last
        child is finished with the parent is also finished with
        (flowchart:~27).

A parent mail can have multiple children, which may be delivered before or
after the parent mail.
